---
title: "Introduction to R 2018 Workshop"
author: 'null'
date: 'null'
output:
  html_document:
    depth: 5
    highlight: pygments
    number_sections: no
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
subtitle: Sharks International
---

#Introduction to R programming Workshop June 03 2018

##What we will cover today
This course is intended for people who are interested in learning the basics of programming using the R statistical programming language. No experience in R or programming is necessary! Whether you are an early career scientist or a professional looking to expand your skill set, this course will open the door to a powerful tool that is quickly becoming the gold standard in the biological sciences and essential for analyzing ecological data. This course will cover the basics of programming in R, including writing code, loading data, data manipulation (subset, min, max, summary tables), data visualization, elementary data analysis (e.g. t-test, anovas), and linear modeling. It will also provide a guide to get you acquainted with the (seemingly endless) resources available online, which will help you continue learning about R.

##Who we are
<center>

![**Dr. Dovi Kacev**  
From ugly duckling  
to competent programmR  
Shark scientist too  
](images/Dovi.png)

![**Lindsay Davidson**
Lindsay wants to get your science R-ipped! She has done the heavy-lifting of learning R and has worked tirelessly on her curls and push-ups to get the dplyr pipes that will change the way you do science. Outside of the gym, she works to inform conservation and management priorities for shark and ray species.](images/Lindsay.png)

![**Dr. Seb Pardo**
As a member of Team Cannonball, Seb will help you program in R efficiently to ensure your code flies! When not going Bayesian, he spends his time working with R code studying a broad range of taxa, from birds, to sharks, rays, and even Atlantic salmon.](images/Seb.png)

![**Dr. Christopher Mull**
Chris (AKA Drsharkbrain) spends his time shreddin’ the gnar and codin’ in R. When he isn’t busy studying shark and ray brains at Simon Fraser University he can be found in the studio, RStudio, laying that sweet RMarkdown. Chris has spent too many years of his life painstakingly crafting the most radical code and exploring the deep mysteries of the shark and ray tree of life. He is sure to help make your functions the object of envy for all in your workspace.](images/Chris.png)

</center>

##Other stuff
For today you will need to have access to a computer with the latest version of [R (v3.3 or greater)](https://www.r-project.org/) and the latest version of [R studio](https://www.rstudio.com/) installed.

#Using R

##R Basics, Objects, Operators
Annote your files by using # before the text. Save you R code with ctrl s. Note your objects will not be saved, just your R script!  

Create an object `a` using the `<-` syntax. Here you are creating an object called `a` that is the sume of 1+1.
```{r R basics}
a <- 1+1
```

What is `a`?
```{r}
a
```

Create a new object called `a` and overwrite the old a - **note** R will not tell you if you overwrite an object.
```{r}
a <- 10
```

What is the new value of `a`?
```{r}
a
```

Objects can be a number, character, string of numbers, matrix, etc. Here, you can create an object with more than one value, the 'c' stands for concatenate.
```{r}
b <- c(10,1,6,0)

char <- "charactertext"

stringofchar <- c("sharks", "rays", "chimaeras")
```

What is stored in your object `b`, `char`, and `stringofchar`?
```{r }
b

char

stringofchar

```

How do you create an object `a` that is two values 1 and 2?
```{r}
a<-c(1,2) #Note to self

a

```

Overwrite `b` to be object `a`
```{r}
b <- a

b
```

You can also define objects using math operators, `*` is multiplication. 
```{r}
a<-3*3

a
```

Use math operators to define a string of value. 
```{r}
a <- c(1+2,10*10,1-1)

a

```

You can also use other operators such as mean, sum, min, and max.
```{r}
x <- mean(a)

y <-sum(a)

y <-max(a)

```

Write out two ways to create an object `z` that is the sum of 1,2,3,4,5,6,7.
```{r}
z<- sum(1,2,3,4,5,6,7,8)

z<- (1+2+3+4+5+6+7+8)

```

If you run the following line of code, what happens? Look at the error message - what does it mean?
```{r, error=TRUE, warning=TRUE, message=TRUE}
a<-3*
```

##Importing
Load up a .csv file called `brain.data.csv`. R works with .csv files as opposed to excel files (.xls or .xlsx).  
In your File Explorer navigate to the folder where your save the `brain.data.csv` file.  
You can set the working directory `setwd` so that R will automatically check that folder when you load up files. If you file is not in the folder that you `setwd` to, you will get an error message. 

Check what the working directory is. Remember, the working directory is the folder on your computer that R will look for files in.  
```{r}
getwd() 

```

Change the working directory (wd) to the folder with the `brain.data.csv` file. **Note** in R the file paths have forward slashes. Such as C:/Workshop
```{r, eval=FALSE}
#setwd("C:/your-file-path/shaRk-workshop/Code") 

```

Use the `read.csv` function to pull your .csv file into R. Here we are creating an object called `data` and define that object as your .csv file
```{r importing data}
data <- read.csv("data/brain.data.csv")
```

Take a look at at your `data` object.
```{r}
data
```

That is a lot to look at..use the `head` function to look at the first 6 rows of your object `data` and `tail` to look at the last six rows.
```{r}
head(data)

tail(data)
```

Look at the column names in your database. **NOTE** having spaces or special characters in your database or column names is not a good idea! Remove them and replace with "." or "_".
```{r}
names(data)
```

Use the `$`to query a column in your object `data`. If you want to look just at the column names in the `data` object, use the `names`function. 
```{r}

names(data)

data$species

data$family

```

The $ acts as a way to query the database, you will get an error is the column name is spelt wrong or doesn't exist. 
```{r, meesage=TRUE, error=TRUE}
data$Species
```

Write out code to query the column called `superorder`:
```{r, eval=FALSE}
data$superorder
```

You may want to know other details about the type of data in your database. What do the following lines of code tell you about your database?
```{r, eval=FALSE}
class(data)
str (data)
dim(data)
length(data)
```

Find the maximum value of brain mass in the dataset from the column brain.mass 
```{r}
max(data$brain.mass)

```

Create an object `a` that is the maximum brain size in `data`.
```{r}
a <-max(data$brain.mass)

```

What is the value of `a`?
```{r}
a
```

Write out the code for finding the smallest and mean brain size in the database.
```{r}
min(data$brain.mass)

mean(data$brain.mass)

```

Create a new column in your database that is called `newcolumn` and is defined as the column `body mass` divided by 1000.
```{r}
data$newcolumn <- data$body.mass/1000

```

Find the number of unique species in the `data` object.   
First, what is the name of the column that has the speies names?  
```{r}
names(data)

```

Second, use the `length` function to tell us how many entries are in the database. Does this tell us the number of unique species names?
```{r}
length(data$species)

```

Use the `unique` function and the `length` function to count the number of unique species names
```{r}
length(unique(data$species))

```

Create a new object that is only the unique family names in `data`.
First, find the column that has the family names by using `names`.
```{r}
names(data)
```

Second, create a new object called uniquefam and define it a the unique names from the family column. 
```{r}

uniquefam <- unique(data$family)
```

Now use the `subset` function to create a new database. Make the database only those entires under the column `family` that are **Rajidae**. Call the new database `rajidaedata`. **NOTE** the `==` syntax.
```{r}
rajidaedata <- subset(data, data$family =="Rajidae")
```

How many entries are in the `rajidaedata` database?
```{r}
dim(rajidaedata)
```

Now create a database that is all families **BUT** rajidae. Use the `!=` syntax. with ! meaning does not equal.
```{r}

norajidae <- subset(data, data$family != "Rajidae")

```

Do this again, but now create a database that is ONLY Rajidae **AND** Myliobatidae. Call the new database `mylioraji`
```{r}

mylioraji <- subset(data, data$family == "Rajidae"| data$family == "Myliobatidae")

```

To get more information on a function, use the ? and then the function name. 
```{r, eval=FALSE}
?subset

```

 The square brackets `[]` are a way to subset vectors or dataframes. The database has two dimenions - rows and columns, and cal be references seperately or together and are seperated by a comma. You can also use the `:` to mean through.
```{r}
data[1,]

data[,1]

data[1,1]

data[1:10,1:10]
```

Create a new object called `noduplicates` and remove rows with duplicated species names removed. Use the ! syntax to specific that the rows cannot be duplicates.
```{r}
noduplicates <- data[!duplicated(data$species),]
```


Create a new database that has only the columns `telecephalon`, `medulla`, and `species`. 
```{r}
newdata <- data[,c("telecephalon", "medulla", "species")]
```

Create a new database that has only has the 1,10,and 30th row.
```{r}

newdata <- data[c(1,10,30),]

```

##Libraries
In R you can load packages where scripts and functions have already been written. To use a library, you first need to install the package. You use the `install.packages` function and specify the package inside the " ". Here we are loading in the `tidyverse` package. Note, you will have to load the library each time you open R but only install the package once.
```{r, eval=FALSE}
install.packages("tidyverse")
```

You will be asked to select a mirror four your download, select your location. Now load the library.

```{r}
library(tidyverse) 
```

If you want to get some more information about the package use the `help` function. 
```{r, eval=FALSE}
help(package = "tidyverse")
```

If you use a package, use the `citation` function to get the proper attibution. 
```{r}
citation("tidyverse")

```


##Summarizing
#dplyr is a package that will make filtering, subsetting, summarizing, and transforming your data much, much easier. It is less verbose that other packages and is simple too implement. 
There are a number of functions within the `dplyr` package such as:  `select`  `filter`  `summarize`  `mutate`  `group_by`
Let's first use the select function.  `select` is a way to choose columns of a database. Here, you can select just one column called `species`, two columns (`species` and `family`) or all columns from `species` to `family`. Use `?select` to get more information.  
```{r}
names(data)

data_spp <- select(data, species)

data_sppfam <- select(data, species, family)

data_sel <- select(data, species:family)

```

Instead of using the `subset` function that we used above. `dplyr` has a similar function called `filter`. The advantage with the `filter` function is that it is less verbose and more powerful. As with `subset` you can use the operators:  `==`  `>`  `<`  `<=`  `!=` and others. You can also use `&` for multiple criteria, `|` for meeting optional criteria (this or this). 
```{r}
head(data)

data_batoids <- filter(data, tribe == "Batoids")

data_bigbrain <- filter(data, brain.mass > mean(brain.mass))

data_reefbigbrain <- filter(data, habitat == "Reef" & brain.mass>= mean(brain.mass))

head(data_reefbigbrain)
```

You can use the %in% to filter based on finding values in a string or another database. Here I create a string of 5 randomly selected species and filter the database based on those 5 species. 
```{r}
fivespp <- data[sample(1:35,5), "species"]

data_5spp <- filter(data, species %in% fivespp) 
```

`Summarise` is a function that will create a new database summarizing your data. `mutate` will add a new column to your database.  
```{r}
summarise(data, length(unique(species)))

mutate(data, length(unique(species)))

```

*Pipes `%>%`. You can perform calculations (summarize, mutate, etc.) on subsets of the data (using `group_by`) and then put that database back together. The `%>%` syntax lets you perform multiple functions. The `group-by` syntax specifies how you want the database split. Above, we counted the number of unique species in the database. Here, we are going to count the number of unique species in each of the superorders and put the results in a new database. 
```{r}

numspp<-data %>% group_by(superorder) %>% summarize(NumSpecies = length(species))


```

Instead of creating a new data.frame with `summarize`, you can add a column using `mutate`
```{r}
data <- data %>% group_by(superorder) %>% mutate(NumSpecies = length(species))
head(data)
```

Create a new database that summarizes `data` and has a new column that counts the number of species that are lecithotrophic or matrotrophic. Use the `names(data)` to see the column names and use the `length` fucntion to count the number of species.  
```{r}
countlm<-data %>% group_by(reproductive.mode) %>% summarize(numspecies = length(species))
```

Create a new database that calculates the mean `brain.mass` of the four superorders but exlude the Carcharhiniformes. 
```{r}
mean<-data %>% group_by(superorder) %>% filter(family != "Carcharhiniformes") %>% summarize(meanbrainmass = mean(brain.mass))
```


##Introduction to ggplot
Install the package `ggplot` and load the library. Also use ?ggplot to see more information about this library. 

```{r ggplot, eval=FALSE}
install.packages ("ggplot2") 
library(ggplot2) 
```

ggplot is an easy way to plot you data. 
```{r}
ggplot(data,aes(x=body.mass,y=brain.mass))+
  geom_point()

```

Colour the points according to whether the species is lecithotrophic or matrotrophic.
```{r}
ggplot(data,aes(x=body.mass,y=brain.mass))+
  geom_point(aes(colour = as.factor(reproductive.mode)))

```

Add a line to the points
```{r}
ggplot(data,aes(x=body.mass,y=brain.mass))+
  geom_point(aes(colour = as.factor(reproductive.mode))) +
  geom_line()

```

Make the lines the same colour as the dots
```{r}
ggplot(data,aes(x=body.mass,y=brain.mass))+
  geom_point(aes(colour= reproductive.mode)) +
  geom_line(aes(colour= reproductive.mode)) 

```

Define your own colours
```{r}
ggplot(data,aes(x=log(body.mass),y=log(brain.mass)))+
  geom_point(aes(colour= reproductive.mode)) +
  geom_line(aes(colour= reproductive.mode))  +
  scale_color_manual(name = "reproductive.mode",
                     values = c("Lecithotrophic" = "red",
                                  "Matrotrophic" = "blue")) +
  xlab("Body mass")+ylab("Brain Mass") +
  theme_classic() 

```


##Introduction to stats
Run a t-test on you data. What is the error message?
```{r, message=TRUE, error=TRUE}
t.test(data$brain.mass~data$species)


```

The error message is telling you that you have more than 2 levels, ie two species in your database. Therefore subset the database to include only 2 species. 
```{r}
mylioraj <-subset(data, data$family =="Rajidae"
                         | data$family == "Myliobatidae")
```

Now run the t.test
```{r}
t.test(mylioraj$telecephalon~mylioraj$family)

```

Run a t-test to see if brain.mass is different between the two reproductive modes.
```{r}
t.test(data$brain.mass~data$reproductive.mode)
plot(data$brain.mass~data$reproductive.mode)

```

Try an anova to test for difference in brain.mass betweem all families. Make sure subject column is a factor, so that it's not treated as a continuous variable. 
```{r}
str(data)

```

Ours is a factor but if it wasn't you would run:
```{r}
data$family <- factor(data$family)

```

Run the anova
```{r}
aov1 <- aov(brain.mass ~ family, data=data) 

```

Check the output. 
```{r}
summary(aov1)
model.tables(aov1, "means")
boxplot(brain.mass ~ family, data=data)

```


Run an anove between family and habitat. 
```{r}
aov2 <- aov(body.mass ~ family*habitat, data=data) 
summary(aov2)
model.tables(aov2, "means")

```


Run an anova
```{r}
aov2 <- aov(body.mass ~ family+habitat, data=data) 
summary(aov2)
model.tables(aov2, "means")


```


Run a linear model with the `lm` function. 
```{r}
lm(telecephalon ~ medulla, data= data)

```

Create a plot that has telecephalon on the y-axis and medulla on the x-axis. 
```{r}
plot(telecephalon ~ medulla, data= data)

```


Plot the data, make it look good, add a title etc. Add a line that shows the model outputs using the `abline` function with the lm nested. 
```{r}
plot.new
plot(telecephalon ~ medulla, data = data, pch=21, 
     bg=c("red","green3","blue")[unclass(data$reproductive.mode)], 
     main="Brain data", xlab="telecephalon", ylab="medulla")

abline(lm(telecephalon ~ medulla , data = data), col="black")


```


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
