---
title: "shaRk-workshop - Further resources"
author: "Sebastian Pardo"
date: "April 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Further resources

One of the most important things to learn as you're beginning to use R is to 
**learn to learn**; that is, to be able to answer (or find the answer) to the endless
questions that you'll have as you dwelve deeper into this programming language and its 
tools.

### R's help files

Every function in R is required to have an associated help file fully explains what the function does, what its inputs are (arguments), and what the outputs are (values). A lot of people are intimidated by these functions, but they are actually really helpful! For example, the help file for the `mean` function:

```{r}
?mean
```

If you want to search for something that is not a function, you can use the `??` to search for a term:

```{r}
??average
```

This will show all the functions from the packages you have installed that mention the search term. The more installed packages, the more potential terms that will be shown.

### CRAN's Introcduction to R 

This is the "official" Introduction to R document developed by the creators of R.
It is pretty comprehensive while using a language that caters more towards programmers:

<https://cran.r-project.org/doc/manuals/R-intro.pdf>

### Swirl

[Swirl](http://swirlstats.com/) is an interactive R package that lets you learn R, in R.
It consists of modules you load and run via the R console: 

```{r eval=FALSE}
install.packages("swirl")
library("swirl")
swirl()
```

### Stack Overflow

I'd a safe bet that if you have a beginner-level R question it has already
been asked online, and it's most likely on [Stack Overflow](http://stack-overflow.com). 
There are over [235,000 questions](https://stackoverflow.com/questions/tagged/r) relating to R on that site alone!

Everybody is welcome to post new questions, however first check that you're not duplicate
a question that has been previously asked/answered. Also, provide a **Minimum Working Example** (MWE)
that reproduces the hurdle you're trying to overcome. Provide a dummy dataset if necessary!

## Google

Lastly, whenever you get an error message you don't understand, googling the e message + function name + CRAN (the R package repository) will usually return questions and answers written by other people about the same or a similar issues.

